Debugging by \textit{lastChange}

### 2010-11-23 papers/ecoop2011/ecoop2011-lastChange.tex
edit section 5 to reorder subsections and change intro

\section{Reproducible Non-deterministic Execution}
We claimed that the only prerequisite for \textit{lastChange} function is bug
reproducibility. A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. When the developer selects a value and
asks for \textit{lastChange}, the developer is expressing the hypothesis that
the value is related to the defect. Therefore the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution. In this section we discuss issues regarding to non-determinism.

@@

\subsection{\textit{lastChange} Result Consistency}
Each time we re-execute a non-deterministic program, 
the details of execution instruction order may change. 
For example, if we record the source code lines every time
a conventional watchpoint hits, the record may differ each time we
re-execute. But \textit{lastChange} does not ask the developer to
consider these single points. Rather it analyzes all of the
change events in an entire execution at the reproduction point where the bug
has already occurred and shows the final result to the developer.

Of course there can be cases where the bug is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventional
breakpoint debugging, implementations of \textit{lastChange} readily
help in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic. Consider the example in Figure~\ref{fig:nondeterministic-values}. 
Assume that at the first exeuction, developers sees that \texttt{a} is \texttt{true} and asks
for the \textit{lastChange} of \texttt{a} at breakpoint. In the re-exeuction, \texttt{a} is \texttt{false},
and \texttt{b} is \texttt{b} is \texttt{true} at the reproduction point. It means that \texttt{lastChange}
result shows line 10, which is a correct answer for this execution but not for the previous one.
When user asks for the \textit{lastChange} on variable \texttt{a}, debugger stores \texttt{a} value and compares
it to the \texttt{a} value in the next executions. So if this value is different, debuggers inform the developer
that the \textit{lastChange} query is made on a different value. In cases that the value is not a primitive but an object,
the properties of the object examined by the developer are stored and compared.

>>>>

\section{Reproducible Non-deterministic Execution}
We claimed that the only prerequisite for \textit{lastChange} function is bug
reproducibility. A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  All modern debuggers in wide use that we know 
about rely on reproducible but non-deterministic execution for simple practical reasons: 
developers must reproduce a bug to study it and modern execution environments 
are not deterministic.

 In this section we discuss non-determinism and \textit{lastChange}. However we want 
to emphasize that when developers have to think about non-determistic execution 
it usually means that bugs are not reproducible. 
The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. Conversely, if the defect depends on order of execution, 
it will likely fail to reproduce on every execution. By relying on reproducibility but 
not requiring deterministic execution, \textit{lastChange} covers the common case 
developers focus on while not adding addition complexity to the implementation.

\subsection{\textit{lastChange} Result Consistency}
Each time we re-execute a non-deterministic program, 
the details of execution instruction order may change. 
For example, if we record the source code lines every time
a conventional watchpoint hits, the record may differ each time we
re-execute. But \textit{lastChange} does not ask the developer to
consider these single points. Rather it analyzes all of the
change events in an entire execution at the reproduction point where the bug
has already occurred and shows the final result to the developer. 
Neither the data gathering nor the analysis require deterministic execution.

When the developer selects a value and
asks for \textit{lastChange}, the developer is expressing the hypothesis that
the value is related to the defect. Since the defect does not depend on
 exeuction order, the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution.
Of course there can be rare cases where the bug is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventional
breakpoint debugging, implementations of \textit{lastChange} readily
help in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic.

Consider the example in Figure~\ref{fig:nondeterministic-values}. 
Assume that at the first exeuction, developers sees that \texttt{a} is \texttt{true} and asks
for the \textit{lastChange} of \texttt{a} at breakpoint. In the re-exeuction, \texttt{a} is \texttt{false},
and \texttt{b} is \texttt{b} is \texttt{true} at the reproduction point. It means that \texttt{lastChange}
result shows line 10, which is a correct answer for this execution but not for the previous one.
When user asks for the \textit{lastChange} on variable \texttt{a}, debugger stores \texttt{a} value and compares
it to the \texttt{a} value in the next executions. So if this value is different, debuggers inform the developer
that the \textit{lastChange} query is made on a different value. In cases that the value is not a primitive but an object,
the properties of the object examined by the developer are stored and compared.

@@
